### 2018-09-11 Manuscripts/cog-psych/revision/cogpsych.tex
Final revisions by Joe. Mostly cleaned up various loose bits. Now off to Ken.

% ------- ABSTRACT PAGE ------- %
\doublespacing
\section*{Abstract}
The ability to generate new concepts and ideas is among the most
fascinating aspects of human cognition, but we do not have a strong
understanding of the cognitive processes and representations underlying concept
generation. In this paper, we study the generation of new categories using the
computational and behavioral toolkit of traditional artificial category
learning. Previous work in this domain has focused on how the statistical
structure of known categories generalizes to generated categories, overlooking
the degree of contrast between the known and generated categories as a factor.
We report three experiments demonstrating that contrast between what is known
and what is created is of fundamental importance for categorization and innovation. We propose two novel approaches to modeling category contrast:
one focused on exemplar dissimilarity and another on the representativeness
heuristic. These new models are compared to two key alternatives. Our
experiments and simulations demonstrate specifically how category contrast
influences creative generation, and how it can interact with other constraints
to produce different types of creations. Our work also serves as a concrete
example of how well-established and highly controlled experimental frameworks
(such as those in the field of category learning) can be used to make progress
in an intriguing yet understudied domain.

\setlength\parindent{0.5in}
{\em Keywords}: categorization; concepts; creativity; generation; computational modeling; exemplar models
\clearpage

@@

Creating and learning new ideas is one of the most fascinating and important
human capabilities. For example, cell phones and smart phones are newly created
categories of objects that young and old adults learned and have become reliant
on in the last few decades. In fact, computational researchers have identified
generating novel objects and objectives as one of the defining, and most
difficult to formalize, characteristics of intelligent life
\citep{lake2017,lehman2011,taylor2016}. Yet, the cognitive mechanisms that
enable people to innovate and learn new innovations are not well understood and
are understudied by scientists. In part, this is due to the inherent difficulty
of designing and conducting experiments that test these capabilities. How would
a scientist devise an experiment that induces a participant to create and learn
new categories that are as interesting as cell phones and smart phones? In this
article, we use computational models and behavioral experiments to investigate
the prior expectations people have when they learn and generate new categories.

Generating and learning new concepts are, however, not altogether different from
the types of behaviors typically studied in cognitive psychology laboratories.
In particular, generating a member of a novel class can be considered a `special
case' application of existing category knowledge
\citep{kemp2014taxonomy,kurtz2015human}. Research in categorization typically
focuses on what properties of a category affect the classification of an
object's category given its features \citep{shepard1961learning,kurtz2013human},
and the prediction of an object's unobserved features given some of its other
features and/or its category \citep{markman2003category}. The generation of
members of a new category consists of inferring {\em all} features for a {\em
novel} category label. Thus, we can make progress formalizing the processes
involved in category generation by extending theories of categorization to this
case. Then, to ensure that these principles reflect basic principles of
categorization and are not just specific to category generation, we can examine
what factors of the generated categories affect the ease at which people learn
the category.

Previous work in category learning and generation has established that people
are highly sensitive to the structural properties of categories, such as
correlations between the features of category members and the relation between
items within the same category and those in different categories
\citep{roschmervis1975,regier2007,shepard1961learning,smith1995creative}.
Inspired by this work, previous research on the topic of category generation has
explored a similar principle: People tend to create new categories that have
similar {\em statistical regularities} as previously learned categories
\citep{jern2013probabilistic,ward1994structured}. Although this is an important
characteristic of generating new categories, it cannot be the only one. Taken to
the extreme, the best ``new'' category in terms of having the same statistical
regularities to other categories would be identical to a known category that is
representative of the domain (and thus, not new at all). Further, contrast has played a critical role studying the spread of new ideas in diverse fields, such as marketing \citep{berger16} and sociology \citep{rogers03}.

@@

We examine two different definitions of how categories should differ from one
another: exemplar dissimilarity or representativeness of the alternative
category. To formalize the first technique, we build on the largely successful
exemplar modeling framework that
\citep{medin1978context,nosofsky1984choice,nosofsky1986attention}. In doing so,
we propose a novel exemplar model of category generation, {\em Producing Alike
  and Contrasting Knowledge using Exemplar Representations} (PACKER),
formalizing how new categories should differ from previous categories. It
embodies the exemplar dissimilarity principle by incorporating a factor that
repulses members of opposite categories from one another. To foreshadow, we find
that this notion of contrast predicts human category learning performance, above
and beyond other category learning models in a novel behavioral experiment. This
demonstrates that contrast plays a central role in other important
categorization tasks, and it is not specific to category generation.

The second hypothesis for how contrast might affect categorization is contrast
as {\em representativeness}. The representativeness heuristic \citep{kahneman73}
states that people make judgments regarding an outcome based on how
representative it is of the evidence given in the current context. The heuristic
is a powerful theoretical construct that has been used to capture a range of
complex patterns of human judgments \citep{kahneman73,tversky74,tversky83},
especially those that deviate from normative theory (as given by a
straightforward application of Bayes' rule). For example, the coinflip sequence
TTHTH is perceived to be more random (due to it being representative of a
coinflip sequence) than TTTTT (despite both having the same probability of
having been generated from a fair coin). Although there is a healthy debate as
to whether perceived randomness {\em really} deviates from normative theory
\citep{griffiths18,hahn09}, the representativeness heuristic remains one of the
main explanations of human judgments \citep{reimers18}. To formalize the
hypothesis that people expect a new category to be representative of the
opposite of the current category, or contrast as {\em representativeness}, we
use the Bayesian formulation of representativeness \citep{tenenbaum01}, which
has been used successfully to capture human performance in color categorization
\citep{abbott16}, image category learning \citep{abbott11}, and language grammar
learning \citep{rafferty10}. We do so by extending the state-of-the-art category
generation model \citep{jern2013probabilistic} and find that human category
generation is best captured as generating representative, rather than probable,
samples.

The outline of the article is as follows. First we describe previous empirical
work on the topic of category generation, as well as the computational
formalizations of their theoretical explanations. We then present two hypotheses
for how contrast might affect categorization and formalize them in computational
models. The first is a novel exemplar model, which is designed to generate
categories that systematically differ from existing categories in the domain.
The second assumes the goal of category generation is to create representative
samples of the opposite of the observed categories. We present two experiments
demonstrating strong and systematic effects of category contrast on concept
generation, and we qualitatively and quantitatively analyze the performance of
each model in capturing human category generation. Afterwards, we find support
for category contrast predicting the ease at which people learn a pair of 
categories in a novel empirical experiment. We conclude with a discussion of the
implications of our results for categorization and creative cognition, and
directions for future work.

@@

\sxlnote{Added thomas98 description here} \jlanote{still need the contrast result from her work.} It is worth pointing out that the
observation of regularities in distributional properties across categories is
not confined to generating whole new categories, but it also occures when inferring an unobserved feature within traditional categorization paradigms. Most notably, \citet{thomas98}
found that when people were asked to generate feature values for exemplars with
a missing feature (specifically circles of different sizes, each with a radial
line of varying orientation), the resulting set of exemplars tended to share the
same feature correlations as exemplars from previously learned categories. [the
transition to the next point isn't great, though]

The broader set of observations made by Ward and colleagues provide a great deal
of insight into the role of prior knowledge in constraining category generation.
Much of the work from this area
\citep[e.g.,][]{smith1993constraining,marsh1999inadvertent} focuses on how
information provided to participants (such as an example of a species generated
by other participants) can drastically diminish difference of a new category
from pre-existing categories. Theoretical accounts of these effects have
primarily been grounded within the categorization literature. For example, the
predominant ``Path of Least Resistance'' account
\citep[see][]{ward1994structured,ward1995s,ward2002role} proposes that, when
generating a new species of animal, people retrieve from memory a known
subcategory of animals (e.g., {\em bird}, {\em dog}, {\em horse}), and simply
change some of the features to make something new. People are thought to change
only features that are not characteristic of the retrieved category (e.g., if
{\em bird} was retrieved, the presence of {\em wings} would not change, but {\em
color} might). This theory incorporates elements of the highly influential
basic-level categories framework \citep{rosch1975cognitive,rosch1976basic}, as
well as the exemplar view \citep{medin1978context,brooks1978nonanalytic}. While
this work is been incredibly useful in providing a conceptual sketch of
generation theories, the hand-drawn responses used in the experiments paradigms
precludes the development of formal approaches that can be used to test the
theories in a fine-grained manner.

\cite{jern2013probabilistic} recently showed that concept generation could be
studied in a more controlled manner through the well-developed methods of an
artificial categorization paradigm \citep[see][for a review]{kurtz2015human}. In
Experiments 3 and 4 of their article, participants were exposed to members of
experimenter-defined categories of "crystals" varying in size, hue, and
saturation. Following a training phase during which the experimenter-defined
categories were learned, participants were asked to generate novel categories of
crystals. In a finding mirroring \cite{ward1994structured},
\cite{jern2013probabilistic} found that participants generated categories with
the same distributional properties as the experimenter-defined categories. For
example, after learning categories with a positive correlation between the
size and saturation features (larger sized crystals were more saturated),
participants generated novel categories with the same positive correlation. This
finding is notable, as it demonstrates that category generation can be studied
in a well-known and highly controlled experimental paradigm.

@@

\section{And Now for Something Different: A Role For Contrast}

Prior work on category generation has explored only one factor: distributional
correspondences between learned and generated categories. As a result most of
the computational, theoretical, and empirical efforts have been focused on
explaining those effects. In this paper, we investigate another important
factor: category contrast. To generate a novel concept, individuals must produce
something that is in some capacity {\em different} from what they already know.
As a consequence, we propose that contrast should be a primary constraint that
underlies people's expectations about how a set of categories are organized: Categories should be different from each other.

Although it is evident that people are {\em capable} of creating new concepts
and categories, it is not entirely clear how new concepts are systematically
made different from what is already known. The hierarchical Bayesian model
developed by \cite{jern2013probabilistic} assumes that differences between
observed and generated categories are only due to random variation. The model
assumes that generated categories are sampled from the same underlying domain
distribution as observed categories, and will thus share a common distributional
structure. The model does not make predictions about the {\em location} of the
category within the domain (the perceptual instantiation of category members).
Under a strict interpretation of their model, given knowledge of a single
category within the domain, the most probable new category to be generated is
located in {\em exactly} the same location and possesses an identical
distributional structure. This is not an issue with their model specifically,
but using a broader class of standard hierarchical Bayesian models without any additional features \citep[e.g.,
][]{griffithsscn07,kemp07}. Many of these models assume that at some point of
the latent generative process the same underlying distribution generates all of
the categories and thus, any differences between categories are due to {\em
noise} and should not be {\em systematic}. If the goal of creating a new
category given others in a domain is generating from the posterior predictive
distribution then the best a standard hierarchical Bayesian model without a
notion of contrast built into the model can do at capturing contrast is to
assume that the new category is placed uniformly at random over stimulus space.
But, this defeats the purpose of a hierarchy as it is ignored when determining a
new category location!\footnote{It is plausible that some hierarchical Bayesian
model could be created that generates categories different from each other.
However, this model would not be a standard application or extension of most
pre-existing hierarchical Bayesian models. The generative process would need to
include a component that promotes contrast.} Note that this does not mean
location information and contrast cannot be captured using hierarchical Bayesian
models. In fact, a model instantiating contrast as representativeness is a hierarchical
Bayesian model. But, it requires an additional factor, which in our case is having a different goal than predicting the next category.

@@

As noted above, the constraint that new concepts should differ from what is
already known has been largely overlooked in previous work. This is no doubt in
part due to the vague definition of what it means for a concept to be
``different'': A generated category may be different from what is already known
in any number of respects. Towards providing a more precise definition of the
role of contrast in creative generation, we formalized contrast in a novel
exemplar model, PACKER ({\em Producing Alike and Contrasting Knowledge using
Exemplar Representations}). PACKER explains category generation as a balance
between two fundamental constraints: The category to be generated should not be
similar to known categories, and exemplars within each category should be
similar to one another. These ideas are implemented within the well-studied
exemplar framework -- the PACKER model is an extension of the influential
Generalized Context Model of categorization
\citep[GCM;][]{nosofsky1984choice,nosofsky1986attention}.

Although, as an exemplar model, one of PACKER's proposals is people represent
categories in terms of a collection of stored exemplars, it is worth emphasizing
that the underlying nature of human category representations is not the primary
concern of our work. The choice to develop PACKER within an exemplar framework
reflects the facts that exemplar models have been throughly evaluated, are
strongly theoretically motivated, and dominate much of the theoretical and
empirical work in categorization. The focus of our work with PACKER concerns the
dual constraints of within- and between-class similarity; it is not difficult to
imagine how such constraints may be instantiated using alternative frameworks
(e.g., \citealp{kurtz2007divergent,love2004sustain,smith2000thirty}; for a
review of categorization models, see \citealp{pothoswills2011}).

@@

PACKER (as well as its name) was in part inspired by earlier work from the
categorization literature \citep{hikada2011packing,stewart2005}. \citet{hidaka2011packing} argued that
natural categories ``pack'' the values of features such that different
categories fill the domain space with distance between one another, while
keeping items within the same category close together. Inspired by this idea,
PACKER proposes that generation is constrained by both similarity to members of
the target category (the category in which a stimulus is being generated) as
well as similarity to members of other categories: the most desirable generation
candidates are similar to members of the target category and not similar to
members of contrast categories. This is achieved by aggregating similarity
across known exemplars differently according to class membership. The aggregated
similarity $a(y,x)$ between generation candidate $y$ and stored exemplars $x$ is
given by:

@@

\jlanote{I'm a bit confused on the theta constraints. it's clear they need to be non-negative, but you don't usually see both needing to be between 0 and 1 but not sum to some value. is this really what we ended up with? Something feels fishy to me with the math.}
PACKER sets $f(x_j)$ depending on exemplar $x_j$'s category membership:
$f(x_j) = \theta_t$ if $x_j$ is a member of the target category, and
$f(x_j) = -\theta_c$ if $x_j$ is a member of a contrast category.
$\theta_t$ and $\theta_c$ are free parameters
($0 \leq (\theta_t,\theta_c) \leq 1$) controlling the trade-off
between within- and between-category similarity. For example, when
$\theta_t = \theta_c = 0.5$, $f(x_j) = 0.5$ for members of the
target category and $f(x_j) = -0.5$ for members of other categories; thus, the
model is likely to generate items that are similar to members of the target
category but are not similar to members of other categories. In this way,
$\theta_t > 0$ with $\theta_c = 0$ produces exclusive consideration
of target-category members, and $\theta_c > 0$ with $\theta_t = 0$
produces exclusive consideration of contrast-category members. The combination
of $\theta_t$ and $\theta_c$ parameters thus specifies a wide
breadth of possible approaches; by fitting it to a dataset, one can describe the
relative roles of between-category contrast and within-category similarity in
generation. See Figure \ref{fig:packer-examples} for an illustration of how
these parameters control the relative influence of within-category similarity
and contrast to other categories when generating a new exemplar.

@@

The probability that a candidate $y$ will be generated is evaluated using an
Exponentiated \citet{luce1977choice} choice rule. Candidates with greater values
of $a(y,x)$ are more likely to be generated than candidates with smaller values:
% 
\begin{equation} p(y \mid x) = \dfrac { \exp \left \{ a \left( y, x
\right) \right \} } { \sum_i{ \exp \left \{ a \left( y_i, x \right)
\right\} } }
\label{eq:packer-choice}
\end{equation}
% 
%where $\theta$ ($\theta \geq 0$) is a free parameter controlling response
%determinism.

It is worth noting that PACKER is only one possible exemplar-based account of
category generation within our proposed framework. That is, PACKER places
specific constraints on the possible values of $f(x_j)$, but other
exemplar-based category generation models with drastically different behavior
can be formalized in this framework by imposing alternative constraints. For
example, as will be discussed in more detail below, PACKER is formally
equivalent to the copy-and-tweak model proposed by \cite{jern2013probabilistic}
when $\theta_c = 0$ and $\theta_t > 0$. Likewise, when
$\theta_t = 0$ and $\theta_c > 0$, PACKER can represent a
contrast-only generation mode, relying exclusively on contrast when generating
new categories. When $f(x_j) < 0$ for all $x_j$ (regardless of class
membership), a ``pure-packing'' approach is yielded, generating items in
unoccupied areas of the domain. Thus, the proposed framework may be used to
describe a wide variety of qualitatively distinct generation strategies.

\subsubsection{Relation Between PACKER and Copy-And-Tweak}
\label{section:copytweak-packer} The PACKER model is similar to the
copy-and-tweak model reported by \cite{jern2013probabilistic}: Both models are
exemplar-based, and both models generate new items according to their similarity
to known members of the target class. However, PACKER diverges from the
copy-and-tweak model by including a contrast mechanism, enabling generation
according to dissimilarity to members of opposing categories. By consequence,
copy-and-tweak can be realized as a parameterization of the PACKER model that is
insensitive to category contrast. Specifically, when $\theta_c = 0$ and
$\theta_t > 0$ (see Figure \ref{fig:packer-examples}, panel B),
$f(x_j) = 0$ for $x_j$ belonging to contrast categories; thus, PACKER is not
influenced by these items, and is mathematically equivalent to a copy-and-tweak
approach.

In this paper, we report simulations using this copy-and-tweak model. This model
fits within the exemplar-based category generation framework defined above, and
is a continuous-dimension adaptation of the model tested by
\cite{jern2013probabilistic}. By formalizing a model family where PACKER and
copy-and-tweak are different parameterizations of models within the same
framework, the comparison between PACKER and copy-and-tweak provides a test of
the explanatory value of the contrast mechanism based on exemplar dissimilarity:
The account provided by copy-and-tweak will only equal that of PACKER if the
contrast mechanism does not offer an advantage (i.e., if $\theta_c > 0$
significantly improves model fits). Note that the purpose of the article is to
explore and formally analyze the role of contrast in categorization and
thus, we leave extending PACKER to incorporate distributional factors \citep[as
explored by][]{jern2013probabilistic} for future work.

@@

\include{methods_results_e3}

>>>>

% ------- ABSTRACT PAGE ------- %
\doublespacing
\section*{Abstract}
The ability to generate new concepts and ideas is among the most fascinating
aspects of human cognition, but we do not have a strong understanding of the
cognitive processes and representations underlying concept generation. In this
paper, we study the generation of new categories using the computational and
behavioral toolkit of traditional artificial category learning. Previous work in
this domain has focused on how the statistical structure of known categories
generalizes to generated categories, overlooking the extent that the contrast
between the known and generated categories is a factor. We report three
experiments demonstrating that contrast between what is known and what is
created is of fundamental importance for categorization. We propose two novel
approaches to modeling category contrast: one focused on exemplar dissimilarity
and another on the representativeness heuristic. Our experiments and
computational analyses demonstrate that both models capture different aspects of
contrast's role in categorization. Our work also serves as a concrete example of
how well-established categorization models can be applied beyond category learning.

\setlength\parindent{0.5in}
{\em Keywords}: categorization; concepts; category learning; generation;
computational modeling; exemplar models; representativeness
\clearpage

@@

Creating and learning new ideas is one of the most fascinating and important
human capabilities. For example, cell phones and smart phones are newly created
categories of objects that young and old adults learned and have become reliant
on in the last few decades. In fact, computational researchers have identified
generating novel objects and objectives as one of the defining, and most
difficult to formalize, characteristics of intelligent life
\citep{lake2017,lehman2011,taylor2016}. Yet, the cognitive mechanisms that
enable people to innovate and learn new innovations are not well understood and
are understudied by scientists. In part, this is due to the inherent difficulty
of designing and conducting experiments that test these capabilities. How would
a scientist devise an experiment that induces a participant to create and learn
new categories that are as interesting as cell phones and smart phones? As a
step towards these goals, in this article we investigate the expectations that
people have when they learn and generate new categories using computational
models and behavioral experiments.

Generating and learning new concepts are, however, not altogether different from
the types of behaviors typically studied in cognitive psychology laboratories.
In particular, generating a member of a novel class can be considered a `special
case' application of existing category knowledge
\citep{kemp2014taxonomy,kurtz2015human}. Research in categorization typically
focuses on what properties of a category affect human learning of an object's
category given its features \citep{shepard1961learning,kurtz2013human}, or the
prediction of an object's unobserved features given some of its other features
and/or its category \citep{markman2003category}. The generation of members of a
new category consists of inferring {\em all} features for a {\em novel} category
label. Thus, we can make progress formalizing the processes involved in category
generation by extending theories of categorization to this case. Then, to ensure
that these principles reflect basic principles of categorization and are not
just specific to category generation, we can examine what factors of the
generated categories affect the ease at which people learn the category.

Previous work in category learning and generation has established that people
are highly sensitive to the structural properties of categories, such as
correlations between the features of category members and the relation between
items within the same category and those in different categories
\citep{roschmervis1975,regier2007,shepard1961learning,smith1995creative}.
Inspired by this work, previous research on the topic of category generation has
explored a similar principle: People tend to create new categories that have
similar {\em statistical regularities} as previously learned categories
\citep{jern2013probabilistic,ward1994structured}. Although this is an important
characteristic of generating new categories, it cannot be the only one. Taken to
the extreme, the best ``new'' category in terms of having the same statistical
regularities to other categories would be identical to a known category that is
representative of the domain (and thus, not new at all). Contrast from other
known categories should play a role. Further, scientists in other fields, such
as marketing \citep{berger16} and sociology \citep{rogers03}, highlight the
critical importance of contrast in the creation of new ideas.

@@

We examine two different definitions of how categories should differ from one
another: exemplar dissimilarity and representativeness of the alternative
category. To formalize the first technique, we build on the largely successful
exemplar modeling framework that
\citep{medin1978context,nosofsky1984choice,nosofsky1986attention}. In doing so,
we propose a novel exemplar model of category generation, {\em Producing Alike
  and Contrasting Knowledge using Exemplar Representations} (PACKER),
formalizing how new categories should differ from previous categories. It
embodies the exemplar dissimilarity principle by incorporating a factor that
repulses members of opposite categories from one another. To foreshadow, we find
that this notion of contrast predicts human category learning performance, above
and beyond other category learning models in a novel behavioral experiment. This
demonstrates that contrast plays a central role in other important
categorization tasks, and it is not specific to category generation.

The second hypothesis for how contrast might affect categorization is contrast
as {\em representativeness}. The representativeness heuristic \citep{kahneman1972subjective}
states that people make judgments regarding an outcome based on how
representative it is of the evidence given in the current context. The heuristic
is a powerful theoretical construct that has been used to capture a range of
complex patterns of human judgments \citep{kahneman73,tversky74,tversky83},
especially those that deviate from normative theory (as given by a
straightforward application of Bayes' rule). For example, the coinflip sequence
TTHTH is perceived to be more random (due to it being representative of a
coinflip sequence) than TTTTT (despite both having the same probability of
having been generated from a fair coin). Although there is a healthy debate as
to whether perceived randomness {\em really} deviates from normative theory
\citep{griffiths18,hahn09}, the representativeness heuristic remains one of the
main explanations of human judgments \citep{reimers18}. To formalize the
hypothesis that people expect a new category to be representative of the
opposite of the current category, or contrast as {\em representativeness}, we
use the Bayesian formulation of representativeness \citep{tenenbaum01}, which
has been used successfully to capture human performance in color categorization
\citep{abbott16}, image category learning \citep{abbott11}, and language grammar
learning \citep{rafferty10}. We do so by extending the state-of-the-art category
generation model \citep{jern2013probabilistic} and find that human category
generation is best captured as generating representative, rather than probable,
samples.

The outline of the article is as follows. First we describe previous
computational formalizations of theories of category generation and empirical
work investigating them. We then present two hypotheses for how contrast might
affect categorization and formalize them in computational models. The first is a
novel exemplar model, which is designed to generate categories that
systematically differ from existing categories in the domain. The second assumes
that the goal of category generation is to create representative samples of the
opposite of the observed categories. We present two experiments demonstrating
strong and systematic effects of category contrast on concept generation, and we
qualitatively and quantitatively analyze the performance of each model in
capturing human category generation. Afterwards, we find support for category
contrast predicting the ease at which people learn a pair of categories in a
novel empirical experiment. We conclude with a discussion of the implications of
our results for categorization and directions for future
work.

@@

The broader set of observations made by Ward and colleagues provide a great deal
of insight into the role of prior knowledge in constraining category generation.
Much of the work from this area
\citep[e.g.,][]{smith1993constraining,marsh1999inadvertent} focuses on how
information provided to participants (such as an example of a species generated
by other participants) can drastically diminish the difference of a new category
from pre-existing categories. Theoretical accounts of these effects have
primarily been grounded within the categorization literature. For example, the
predominant ``Path of Least Resistance'' account
\citep[see][]{ward1994structured,ward1995s,ward2002role} proposes that, when
generating a new species of animal, people retrieve from memory a known
subcategory of animals (e.g., {\em bird}, {\em dog}, {\em horse}), and simply
change some of the features to make something new. People are thought to change
only features that are not characteristic of the retrieved category (e.g., if
{\em bird} was retrieved, the presence of {\em wings} would not change, but {\em
  color} might). This theory incorporates elements of the highly influential
basic-level categories framework \citep{rosch1975cognitive,rosch1976basic}, as
well as the exemplar view \citep{medin1978context,brooks1978nonanalytic}. While
this work has been incredibly useful in providing a conceptual sketch of
generation theories, its qualitative nature and the hand-drawn responses used in
the experiments paradigms precludes the development of formal approaches that
can be used to test them in a fine-grained manner.

\cite{jern2013probabilistic} recently showed that concept generation could be
studied in a more controlled manner through the well-developed methods of an
artificial categorization paradigm \citep[see][for a review]{kurtz2015human}. In
Experiments 3 and 4 of their article, participants were exposed to members of
experimenter-defined categories of "crystals" varying in size, hue, and
saturation. Following a training phase during which the experimenter-defined
categories were learned, participants were asked to generate novel categories of
crystals. In a finding mirroring \cite{ward1994structured},
\cite{jern2013probabilistic} found that participants generated categories with
the same distributional properties as the experimenter-defined categories. For
example, after learning categories with a positive correlation between the size
and saturation features (larger sized crystals were more saturated),
participants generated novel categories with the same positive correlation. By
replicating \cite{ward1994structured}, they demonstrated that category generation
can be studied in a well-known and highly controlled experimental paradigm.

@@

It is worth pointing out that the early observation of regularities in
distributional properties across categories is not confined to the work of Ward
and colleagues. Most notably, \citet{thomas98} trained participants to classify
circles of different sizes, each with a radial line of varying orientation, into
two categories. During a surprise prediction phase, participants were asked to
generate a value for a missing feature for a certain category exemplar given a
value on the other feature. Their results revealed that exemplars tended to
share the same feature correlations as exemplars from previously learned
categories. Interestingly, this effect was not consistently observed when the
given values of a feature fell outside the range of learned values for that
particular category. For instance, if the learned Category A exemplars were
generally 8 cm to 10 cm in radius and negatively correlated with the angle of
the radial line, participants generally produced radial line orientations that
were also negatively correlated with the exemplar size when the given size was
between 8 cm and 10 cm. However, when the given size fell outside that range for
that category (e.g., 14 cm for a Category A exemplar), some participants
consistently produced radial line orientations that were not negatively
correlated with the size and that also placed the exemplars in a significantly
different location in the feature space. These results suggest that when tasked
to produce exemplars that fall outside what was previously learned, people show
some tendency to generate exemplars that are different (both distributionally
and spatially) to what was learned. We explore this idea of contrast
more deeply in the following section.

\section{And Now for Something (Completely) Different: The Role of Contrast}

Prior work on category generation has explored only one factor: distributional
correspondences between learned and generated categories. As a result, most of
the computational, theoretical, and empirical efforts have been focused on
explaining those effects. In this paper, we investigate another important
factor: category contrast. To generate a novel concept, individuals must produce
something that is in some capacity {\em different} from what they already know.
As a consequence, we propose that contrast should be a primary constraint that
underlies people's expectations about how a set of categories are organized:
Categories should be different from each other.

Although it is evident that people are {\em capable} of creating new concepts
and categories, it is not entirely clear how new concepts are systematically
made different from what is already known. The hierarchical Bayesian model
developed by \cite{jern2013probabilistic} assumes that differences between
observed and generated categories are only due to random variation. The model
assumes that generated categories are sampled from the same underlying domain
distribution as observed categories, and will thus share a common distributional
structure. The model does not make predictions about the {\em location} of the
category within the domain (the perceptual instantiation of category members).
Under a strict interpretation of their model, given knowledge of a single
category within the domain, the most probable new category to be generated is
located in {\em exactly} the same location and possesses an identical
distributional structure. This is not an issue with their model specifically,
but using a broader class of standard hierarchical Bayesian models without any
additional features \citep[e.g., ][]{griffithsscn07,kemp07}. Many of these
models assume that at some point of the latent generative process the same
underlying distribution generates all of the categories and thus, any
differences between categories are due to {\em noise} and should not be {\em
  systematic}. If the goal of creating a new category given others in a domain
is generating from the posterior predictive distribution then the best a
standard hierarchical Bayesian model without a notion of contrast built into the
model can do at capturing contrast is to assume that the new category is placed
uniformly at random over stimulus space. But, this defeats the purpose of a
hierarchy as it is ignored when determining a new category location!\footnote{It
  is plausible that some hierarchical Bayesian models could be created that
  generates categories different from each other. However, this model would not
  be a standard application or extension of most pre-existing hierarchical
  Bayesian models. The generative process would need to include a component that
  presumes contrast.} Note that this does not mean location information and
contrast cannot be captured using hierarchical Bayesian models. In fact, a model
instantiating contrast as representativeness is a hierarchical Bayesian model.
But, it requires an additional factor, which in our case is having a different
goal than predicting the next category.

@@

As noted above, the constraint that new concepts should differ from what is
already known has been largely overlooked in previous work. This is no doubt in
part due to the vague definition of what it means for a concept to be
``different'': A generated category may be different from what is already known
in any number of respects. Towards providing a more precise definition of the
role of contrast in generation, we formalized contrast in a novel exemplar
model, PACKER ({\em Producing Alike and Contrasting Knowledge using Exemplar
  Representations}). PACKER explains category generation as a balance between
two fundamental constraints: The category to be generated should not be similar
to known categories, and exemplars within each category should be similar to one
another. These ideas are implemented within the well-studied exemplar framework
-- the PACKER model is an extension of the influential Generalized Context Model
of categorization \citep[GCM;][]{nosofsky1984choice,nosofsky1986attention}.

Although, as an exemplar model, one of PACKER's proposals is people represent
categories in terms of a collection of stored exemplars, we did not pick it
assuming it is the correct model of human categorization. The choice to develop
PACKER within an exemplar framework reflects the facts that exemplar models have
been throughly evaluated, are strongly theoretically motivated, and dominate
much of the theoretical and empirical work in categorization. The focus of our
work with PACKER concerns the dual constraints of within- and between-class
similarity; it is not difficult to imagine how such constraints may be
instantiated using alternative frameworks (e.g.,
\citealp{kurtz2007divergent,love2004sustain,smith2000thirty}; for a review of
categorization models, see \citealp{pothoswills2011}).

@@

PACKER (as well as its name) was in part inspired by earlier work from the
categorization literature \citep{hidaka2011packing,stewart2005}. \citet{hidaka2011packing} argued that
natural categories ``pack'' the values of features such that different
categories fill the domain space with distance between one another, while
keeping items within the same category close together. Inspired by this idea,
PACKER proposes that generation is constrained by both similarity to members of
the target category (the category in which a stimulus is being generated) as
well as similarity to members of other categories: the most desirable generation
candidates are similar to members of the target category and not similar to
members of contrast categories. This is achieved by aggregating similarity
across known exemplars differently according to class membership. The aggregated
similarity $a(y,x)$ between generation candidate $y$ and stored exemplars $x$ is
given by:

@@

PACKER sets $f(x_j)$ depending on exemplar $x_j$'s category membership:
$f(x_j) = \theta_t$ if $x_j$ is a member of the target category, and
$f(x_j) = -\theta_c$ if $x_j$ is a member of a contrast category. $\theta_t$ and
$\theta_c$ are free parameters ($0 \leq \theta_t,\theta_c$) controlling the
trade-off between within- and between-category similarity. For example, when
$\theta_t = \theta_c = 0.5$, $f(x_j) = 0.5$ for members of the target category
and $f(x_j) = -0.5$ for members of other categories; thus, the model is likely
to generate items that are similar to members of the target category but are not
similar to members of other categories. In this way, $\theta_t > 0$ with
$\theta_c = 0$ produces exclusive consideration of target-category members, and
$\theta_c > 0$ with $\theta_t = 0$ produces exclusive consideration of
contrast-category members. The combination of $\theta_t$ and $\theta_c$
parameters thus specifies a wide breadth of possible approaches; by fitting it
to a dataset, one can describe the relative roles of between-category contrast
and within-category similarity in generation. See Figure
\ref{fig:packer-examples} for an illustration of how these parameters control
the relative influence of within-category similarity and contrast to other
categories when generating a new exemplar.

@@

The probability that a candidate $y$ will be generated is evaluated using an
Exponentiated \citet{luce1977choice} choice rule. Candidates with greater values
of $a(y,x)$ are more likely to be generated than candidates with smaller values:
% 
\begin{equation} p(y \mid x) = \dfrac { \exp \left \{ a \left( y, x
\right) \right \} } { \sum_i{ \exp \left \{ a \left( y_i, x \right)
\right\} } }
\label{eq:packer-choice}
\end{equation}
% 
%where $\theta$ ($\theta \geq 0$) is a free parameter controlling response
%determinism.
It is worth noting that PACKER is only one possible exemplar-based account of
category generation within our proposed framework. That is, PACKER places
specific constraints on the possible values of $f(x_j)$, but other
exemplar-based category generation models with drastically different behavior
can be formalized in this framework by imposing alternative constraints. For
example, as will be discussed in more detail below, PACKER is formally
equivalent to the copy-and-tweak model proposed by \cite{jern2013probabilistic}
when $\theta_c = 0$ and $\theta_t > 0$. Likewise, when
$\theta_t = 0$ and $\theta_c > 0$, PACKER can represent a
contrast-only generation mode, relying exclusively on contrast when generating
new categories. When $f(x_j) < 0$ for all $x_j$ (regardless of class
membership), a ``pure-packing'' approach is yielded, generating items in
unoccupied areas of the domain. Thus, the proposed framework may be used to
describe a wide variety of qualitatively distinct generation strategies.

\subsubsection{Relation Between PACKER and Copy-And-Tweak}
\label{section:copytweak-packer} The PACKER model is similar to the
copy-and-tweak model reported by \cite{jern2013probabilistic}: Both models are
exemplar-based, and both models generate new items according to their similarity
to known members of the target class. However, PACKER diverges from the
copy-and-tweak model by including a contrast mechanism, enabling generation
according to dissimilarity to members of opposing categories. As a consequence,
copy-and-tweak can be realized as a parameterization of the PACKER model that is
insensitive to category contrast. Specifically, when $\theta_c = 0$ and
$\theta_t > 0$ (see Figure \ref{fig:packer-examples}, panel B), $f(x_j) = 0$ for
$x_j$ belonging to contrast categories; thus, PACKER is not influenced by these
items, and is mathematically equivalent to a copy-and-tweak approach.

In this paper, we report simulations using this copy-and-tweak model. This model
fits within the exemplar-based category generation framework defined above, and
is a continuous-dimension adaptation of the model tested by
\cite{jern2013probabilistic}. By formalizing a model family where PACKER and
copy-and-tweak are different parameterizations of models within the same
framework, the comparison between PACKER and copy-and-tweak provides a test of
the explanatory value of the contrast mechanism based on exemplar dissimilarity:
The account provided by copy-and-tweak will only equal that of PACKER if the
contrast mechanism does not offer an advantage (i.e., if $\theta_c > 0$
significantly improves model fits). Note that the purpose of the article is to
explore and formally analyze the role of contrast in categorization and thus, we
leave extending PACKER to incorporate distributional factors for future work.

@@

%Experiment 3 starts here
\section{Experiment 3}

Experiments 1 and 2 clearly establish the importance of contrast in category
generation. Follow-up model-based analyses illustrated that both contrast models
account for participant performance better than models that do not take into
account contrast. We also found more support for contrast as representativeness
hypothesis over contrast as exemplar dissimilarity hypothesis.
